Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-06 Manuscript/manuscript.Rmd
manuscript edits

In this study we developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. We provide direct comparisons of cell line specific random forest models built of a variety of different data dependent characteristics, including 4 contact matrix resolutions ($\{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$), 3 predictor-types ($\{\textit{OC}, \textit{OP}, \textit{Distance}\}$), and 4 re-sampling techniques ($\{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$), for the GM12878 and K562 cell lines.

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthew's Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, suggesting that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistenly better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing models.

@@

We also observed that, while the best set of data characteristics were consistent across resolutions, performances of classifiers built on said characteristics tended to decline as resolution decreased. For both cell lines, top performance was achieved at the 10 kb data resolution, while the worst performance was obtained at the 50 kb and 100 kb resolution for the GM12878 and K562 cell line respectively. This result can be attributed to one of two factors. First, at higher resolutions, by model construction genomic bins were smaller, Thereby resulting in more accurate profiles for the genomic annotations. Secondly, because we are mainly concerned with TAD boundaries, and not the heirarchical structure of domains as a whole, the sparcity of contact matrices at higher resolutions does not impair boundary prediction. Conversely, lower resolutions imply larger bin windows and lead to less specificity in boundary location on the linear genome, resulting in worse performances.

>>>>

In this study we developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. We provide direct comparisons of cell line specific random forest models built on a variety of different data dependent characteristics, including 4 contact matrix resolutions ($\{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$), 3 predictor-types ($\{\textit{OC}, \textit{OP}, \textit{Distance}\}$), and 4 re-sampling techniques ($\{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$), for both the GM12878 and K562 cell lines.

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthew's Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, implying that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistenly better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing model.

@@

We also observed that, while the best set of data characteristics were consistent across resolutions, performances of classifiers built on said characteristics tended to decline as resolution decreased. For both cell lines, top performance was achieved at the 10 kb data resolution, while the worst performance was obtained at the 50 kb and 100 kb resolution for the GM12878 and K562 cell line respectively. This suggests that TAD boundary prediction is intimately related to the resolution that boundaries are called on. This result can be attributed to one of two factors in the scheme of our framework. First, at higher resolutions, by model construction genomic bins were smaller, thereby resulting in more accurate profiles for the genomic annotations. Secondly, because we are mainly concerned with TAD boundaries, and not the heirarchical structure of domains as a whole, the sparcity of contact matrices at higher resolutions does not impair boundary prediction. Conversely, lower resolutions imply larger bin windows and lead to less specificity in boundary location on the linear genome, resulting in worse performances. Given the trend in performances across resolution, future research is necessary in determining the optimal resolution to use when predicting TAD boundaries.

Given the formulation of our framework, we were not able to make direct comparisons between cell lines. This was due to the fact that we opted to use cell line specific annotations, and there existed some annotations that were available on the K562 cell line that were not available to GM12878, and vice versa. However, we were able to confirm that final models built on the optimal set of data characteristics identified consistent and well known proteins related to the 3D structure of chromatin, for both cell lines. Our results indicated that the zinc finger protein CTCF, as well components of the cohesin complex, SMC3 and RAD21, were found to be the key molecular drivers most predictive of TAD boundary formation, consistent across cell line and resolution. In addition, models derived from our framework recognized known mediaters of long range enhancer-promoter interactions, the ZNF143 and YY1 transcription factors, as being highly predictive of TAD boundaries.

In summary, we have shown that TAD boundary prediction is a multi-faceted problem composed of many data irregularities and characteristics that previous model have failed to address. We have shown that our proposed framework suitably addresses each of these components, while also providing the optimal considerations to make in an effort to build consensus classifiers used to predict TAD boundary formation from functional genomic annotations. Given the stability of our results, this opens up the possibility of widening our framework to include a broader set of classifiers, cell lines, and Hi-C contact matix resolutions that will inevitably assist us in gaining a better understanding of the mechanisms associated with the architecture of the human genome.
